---
title: "OBIS dataset licenses"
author: "Pieter Provoost"
date: "`r Sys.Date()`"
editor_options: 
  chunk_output_type: console
format:
  html:
    toc: true
    toc-location: left
    toc-expand: 2
    grid:
      sidebar-width: 300px
      body-width: 1100px
      margin-width: 20px
      gutter-width: 1.5rem
---

This is an overview of data licenses applied to OBIS datasets. In particular, it shows the distribution of public domain ([CC0](https://creativecommons.org/public-domain/cc0/)) and open licenses ([CC BY](https://creativecommons.org/licenses/by/4.0/), [CC BY-SA](https://creativecommons.org/licenses/by-sa/4.0/)), versus more restrictive licenses (such as [CC BY-NC](https://creativecommons.org/licenses/by-nc/4.0/)) across datasets and records. Creative Commons licenses other than CC0, CC BY, and CC BY-SA, are considered non-free because they impose restrictions on commercial and derivative use. The former licenses however, are conformant with the Open Knowledge Foundation's [Open Definition](https://opendefinition.org/).

Besides Creative Commons licenses, OBIS datasets also use other licenses such as:

- [Open Government Licence - Canada (OGL-C)](https://open.canada.ca/en/open-government-licence-canada)
- [Open Data Commons Attribution License (ODC-By)](https://opendatacommons.org/licenses/by/1-0/)
- [ICES Data Policy](https://www.ices.dk/data/guidelines-and-policy/pages/ices-data-policy.aspx)

For the purpose of this analysis these licenses will be considered open, but the ICES Data Policy requires closer investigation as it mentions some exceptions on the use of CC BY.

# Dataset metadata

Dataset metadata including licenses can be obtained from the OBIS API using the robis R package.

```{r message=FALSE}
library(dplyr)
library(ggplot2)
library(robis)
library(scales)
library(knitr)
library(stringr)
library(tidyr)

report_theme <- theme(
    axis.text = element_text(colour = "black"),
    axis.line = element_line(linewidth = rel(1)),
    axis.line.y = element_blank(),
    axis.ticks = element_line(), axis.ticks.y = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank(),
    plot.background = element_blank(),
    panel.background = element_rect(fill = "#f5f5f5", colour = NA),
    legend.background = element_blank()
  )

categories <- c("Public domain", "Open", "Non-free")
```

```{r message=FALSE}
#| cache: true
full_datasets <- dataset()
```

```{r}
datasets <- full_datasets %>% 
  select(id, records, intellectualrights, node_name, title) %>% 
  mutate(
    license = case_when(
      str_detect(intellectualrights, "CC-BY\\)?[- ]4.0") ~ "CC BY",
      str_detect(intellectualrights, "creativecommons.org/licenses/by/4.0") ~ "CC BY",
      str_detect(intellectualrights, "CC-BY-?NC\\)?[- ]4.0") ~ "CC BY-NC",
      str_detect(intellectualrights, "CC[- ]BY[- ]SA") ~ "CC BY-SA",
      str_detect(intellectualrights, "CC[- ]BY[- ]NC[- ]SA") ~ "CC BY-NC-SA",
      str_detect(intellectualrights, "CC[- ]BY[- ]NC[- ]ND") ~ "CC BY-NC-ND",
      str_detect(intellectualrights, "CC[- ]BY[- ]ND") ~ "CC BY-ND",
      str_detect(intellectualrights, "(?i)CC0") ~ "CC0",
      str_detect(intellectualrights, "waived all rights") ~ "CC0",
      str_detect(intellectualrights, "creativecommons.org/publicdomain/zero/1\\.0") ~ "CC0",
      str_detect(intellectualrights, "Open Data Commons Attribution") ~ "Open Data Commons Attribution",
      str_detect(intellectualrights, "(?i)unrestricted") ~ "CC0",
      str_detect(intellectualrights, "(?i)^restricted") ~ "Restricted",
      str_detect(intellectualrights, "remain property of") ~ "Restricted",
      str_detect(intellectualrights, "data.gc.ca/eng/open-government-licence-canada") ~ "Open Government Licence - Canada",
      str_detect(intellectualrights, "ICES Data Policy") ~ "ICES Data Policy",
      TRUE ~ "Unknown"
    )
  ) %>% mutate(
    intellectualrights = gsub("\\\n", " ", intellectualrights),    
    category = case_when(
      license == "CC0" ~ "Public domain",
      license == "CC BY" ~ "Open",
      license == "CC BY-SA" ~ "Open",
      license == "Open Data Commons Attribution" ~ "Open",
      license == "Open Government Licence - Canada" ~ "Open",
      license == "ICES Data Policy" ~ "Open",
      TRUE ~ "Non-free"
    )
  )
```

# Records and datasets by license

```{r}
licenses <- datasets %>% 
  group_by(license, intellectualrights, category) %>%
  summarise(datasets = n(), records = sum(records), .groups = "drop") %>%
  arrange(desc(records))

kable(licenses %>% head(10))
```

## Number of records

```{r}
#| fig-width: 12
#| fig-height: 7
#| fig.fullwidth: true
license_stats <- licenses %>% 
  group_by(license, category) %>% 
  summarize(records = sum(records), datasets = sum(datasets), .groups = "drop") %>% 
  arrange(-records) %>% 
  mutate(
    license = factor(license, levels = rev(reorder(license, records))),
    category = factor(category, levels = categories),
    records_percentage = records / sum(records) * 100,
    datasets_percentage = datasets / sum(datasets) * 100
  )

ggplot(data = license_stats) +
  geom_bar(aes(y = license, x = records, fill = category), stat = "identity") +
  geom_text(aes(y = license, x = records, label = paste0(round(records_percentage, 1), "%")), vjust = 0.5, hjust = -0.2) +
  ggtitle("Licenses by number of records") +
  scale_fill_manual(values = c("#7ABD7E", "#FFB54C", "#FF6961")) +
  scale_x_continuous(
    expand = expansion(mult = c(0, 0.1)),
    labels = label_comma()
  ) +
  report_theme
```

## Number of datasets

```{r}
#| fig-width: 12
#| fig-height: 7
#| fig.fullwidth: true
ggplot(data = license_stats) +
  geom_bar(aes(y = license, x = datasets, fill = category), stat = "identity") +
  geom_text(aes(y = license, x = datasets, label = paste0(round(datasets_percentage, 1), "%")), vjust = 0.5, hjust = -0.2) +
  ggtitle("Licenses by number of datasets") +
  scale_fill_manual(values = c("#7ABD7E", "#FFB54C", "#FF6961")) +
  scale_x_continuous(
    expand = expansion(mult = c(0, 0.1)),
    labels = label_comma()
  ) +
  report_theme
```

# Records and datasets by node and license

## Number of records

```{r}
#| fig-width: 12
#| fig-height: 9
#| fig.fullwidth: true
node_license_stats <- datasets %>%
  separate_rows(node_name, sep = ",") %>% 
  group_by(node_name, category) %>%
  summarise(datasets = n(), records = sum(records), .groups = "drop") %>% 
  mutate(
    node_name = factor(node_name, levels = rev(sort(unique(node_name)))),
    category = factor(category, levels = categories)
  )

ggplot(data = node_license_stats) +
  geom_bar(aes(y = node_name, x = records, fill = category), stat = "identity") +
  scale_fill_manual(values = c("#7ABD7E", "#FFB54C", "#FF6961")) +
  scale_x_continuous(
    labels = label_comma()
  ) +
  ylab("node") +
  ggtitle("Number of records by node and license") +
  report_theme
```

## Number of datasets

```{r}
#| fig-width: 12
#| fig-height: 9
#| fig.fullwidth: true
ggplot(data = node_license_stats) +
  geom_bar(aes(y = node_name, x = datasets, fill = category), stat = "identity") +
  scale_fill_manual(values = c("#7ABD7E", "#FFB54C", "#FF6961")) +
  scale_x_continuous(
    labels = label_comma()
  ) +
  ylab("node") +
  ggtitle("Number of datasets by node and license") +
  report_theme
```